---
title: "Zero‑Dose Prevalence in Ethiopia (EDHS 2011–2019)"
author: "Your Name"
date: "2026-07-03"
format: html
editor: source
---

```{r setup, include=FALSE}
#| echo: false
#| warning: false
#| message: false

# Load all required packages
suppressMessages({
  library(tidyverse)
  library(survey)
  library(gtsummary)
  library(flextable)
  library(officer)
})

# Create output directories
dir.create("data/clean/table",   showWarnings = FALSE, recursive = TRUE)
dir.create("data/clean/figures", showWarnings = FALSE, recursive = TRUE)

# knitr options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Load cleaned data

We load the child‑level and cluster‑level RDS files prepared earlier.

```{r read-data}
child   <- readRDS("data/clean/edhs_zero_dose_child_prewored_2026-07-02.rds")
cluster <- readRDS("data/clean/edhs_zero_dose_cluster_prewored_2026-07-02.rds")

# Helper for weighted mean
wtd_mean <- function(x, w) {
  sum(x * w, na.rm = TRUE) / sum(w[!is.na(x)], na.rm = TRUE)
}
```

---

## 2. Descriptive statistics (unweighted and weighted)

We compute four core tables exactly as in your original script and write them to CSVs.

```{r desc-calc}
# Table 1: Round summary
tab_round <- child %>%
  filter(!is.na(zero_dose)) %>%
  group_by(survey_year) %>%
  summarise(
    n_clusters      = n_distinct(cluster),
    n_children      = n(),
    zero_dose_unwtd = mean(zero_dose) * 100,
    zero_dose_wtd   = wtd_mean(zero_dose, wt) * 100,
    .groups = "drop"
  )

# Table 2: Subgroup prevalence
by_group <- function(data, group_var) {
  data %>%
    filter(!is.na(zero_dose), !is.na(.data[[group_var]])) %>%
    group_by(survey_year, .data[[group_var]]) %>%
    summarise(
      n_children = n(),
      zero_dose_wtd = wtd_mean(zero_dose, wt) * 100,
      .groups = "drop"
    ) %>%
    rename(group = 2) %>%
    mutate(group = as.character(group), variable = group_var, .before = 1)
}

tab_subgroup <- bind_rows(
  by_group(child, "residence"),
  by_group(child, "wealth_q"),
  by_group(child, "mother_educ"),
  by_group(child, "child_female")
)

# Table 3: Region prevalence (from cluster file)
tab_region <- cluster %>%
  filter(n_children > 0, !is.na(dhs_region)) %>%
  group_by(survey_year, dhs_region) %>%
  summarise(
    n_clusters  = n(),
    n_children  = sum(n_children),
    n_zero_dose = sum(n_zero_dose),
    prevalence  = n_zero_dose / n_children * 100,
    .groups = "drop"
  ) %>%
  arrange(dhs_region, survey_year)

# Table 4: Cluster summary
tab_cluster_summary <- cluster %>%
  group_by(survey_year) %>%
  summarise(
    n_clusters       = n(),
    n_urban          = sum(urban_rural == "urban", na.rm = TRUE),
    n_rural          = sum(urban_rural == "rural", na.rm = TRUE),
    median_children  = median(n_children),
    clusters_zero_n  = sum(n_children == 0),
    .groups = "drop")

# Write CSVs
write_csv(tab_round,            "data/clean/table/tab1_round_summary.csv")
write_csv(tab_subgroup,         "data/clean/table/tab2_subgroup_prevalence.csv")
write_csv(tab_region,           "data/clean/table/tab3_region_prevalence.csv")
write_csv(tab_cluster_summary,  "data/clean/table/tab4_cluster_summary.csv")
```

---

## 3. Design‑based direct estimates (survey‑weighted)

We prefix PSU and strata with `survey_year` to create unique identifiers across rounds, then use `svydesign` and `svyby` to obtain national and regional estimates with 95% CIs.

```{r direct-calc}
child_design <- child %>%
  filter(!is.na(zero_dose)) %>%
  mutate(
    psu_id    = paste(survey_year, psu),
    strata_id = paste(survey_year, strata)
  )

des <- svydesign(
  ids = ~psu_id,
  strata = ~strata_id,
  weights = ~wt,
  data = child_design,
  nest = TRUE
)

# National
national <- svyby(~zero_dose, ~survey_year, des, svymean,
                  vartype = c("se", "ci"), na.rm = TRUE) %>%
  as_tibble() %>%
  transmute(
    survey_year,
    prevalence_pct = round(zero_dose * 100, 1),
    se_pct         = round(se * 100, 2),
    ci_low         = round(ci_l * 100, 1),
    ci_high        = round(ci_u * 100, 1)
  )

# By region
by_region <- svyby(~zero_dose, ~survey_year + dhs_region, des, svymean,
                   vartype = c("se", "ci"), na.rm = TRUE) %>%
  as_tibble() %>%
  transmute(
    survey_year, dhs_region,
    prevalence_pct = round(zero_dose * 100, 1),
    se_pct         = round(se * 100, 2),
    ci_low         = round(pmax(ci_l, 0) * 100, 1),
    ci_high        = round(pmin(ci_u, 1) * 100, 1)
  ) %>%
  left_join(
    child_design %>% count(survey_year, dhs_region, name = "n_children"),
    by = c("survey_year", "dhs_region")
  ) %>%
  left_join(
    child_design %>% distinct(survey_year, dhs_region, psu_id) %>%
      count(survey_year, dhs_region, name = "n_clusters"),
    by = c("survey_year", "dhs_region")
  ) %>%
  arrange(dhs_region, survey_year)

write_csv(national,  "data/clean/table/direct_est_national_by_round.csv")
write_csv(by_region, "data/clean/table/direct_est_region_by_round.csv")
```

---

## 4. Figures

We generate the same two figures and save them as PNGs.

```{r figures}
# Figure 1: National trend
p1 <- ggplot(national, aes(x = factor(survey_year), y = prevalence_pct)) +
  geom_col(fill = "#2b8cbe", width = 0.55) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15, linewidth = 0.6) +
  geom_text(aes(label = paste0(prevalence_pct, "%")), vjust = -2.6, size = 4.2, fontface = "bold") +
  labs(
    title = "Zero-dose (no DTP1) prevalence, Ethiopia",
    subtitle = "Design-based direct estimate; children 12-23 months\nError bars are 95% CI (survey-weighted)",
    x = "Survey round", y = "Zero-dose prevalence (%)"
  ) +
  ylim(0, max(national$ci_high) + 5) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())

ggsave("data/clean/figures/direct_est_national_trend.png", p1,
       width = 8, height = 6, dpi = 200, bg = "white")

# Figure 2: 2019 regional estimates
region_tab_2019 <- by_region %>%
  filter(survey_year == 2019) %>%
  mutate(region_label = paste0(dhs_region, " (", n_clusters, " clusters)")) %>%
  arrange(prevalence_pct) %>%
  mutate(region_label = factor(region_label, levels = region_label))

p2 <- ggplot(region_tab_2019, aes(x = region_label, y = prevalence_pct)) +
  geom_col(fill = "#fd8d3c", width = 0.65) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.25) +
  coord_flip() +
  labs(
    title = "Zero-dose prevalence by region, EDHS 2019",
    subtitle = "Direct estimate with 95% CI; cluster count shown",
    x = NULL, y = "Zero-dose prevalence (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())

ggsave("data/clean/figures/direct_est_region_2019_ci.png", p2,
       width = 10, height = 6.5, dpi = 200, bg = "white")
```

---

## 5. gtsummary – Baseline characteristics

Here we showcase `{gtsummary}` by creating a demographic summary of the child sample, stratified by survey year.

```{r gtsummary-table}
# Convert relevant columns to labelled factors for a nicer display
child_gtsum <- child %>%
  mutate(
    survey_year = factor(survey_year),
    residence   = factor(residence),
    wealth_q    = factor(wealth_q, levels = c("poorest", "poorer", "middle", "richer", "richest")),
    mother_educ = factor(mother_educ, levels = c("no education", "primary", "secondary+")),
    child_female = factor(child_female, labels = c("Male", "Female"))
  )

tbl_baseline <- child_gtsum %>%
  select(survey_year, residence, wealth_q, mother_educ, child_female) %>%
  tbl_summary(
    by = survey_year,
    statistic = list(all_categorical() ~ "{n} ({p}%)"),
    digits = all_categorical() ~ c(0, 1),
    label = list(
      residence   = "Residence",
      wealth_q    = "Wealth quintile",
      mother_educ = "Mother's education",
      child_female = "Child sex"
    ),
    missing = "no"
  ) %>%
  add_overall() %>%
  modify_header(label = "**Characteristic**") %>%
  modify_footnote(all_stat_cols() ~ "n (unweighted %)")

# Convert to flextable for later use in Word report
ft_baseline <- tbl_baseline %>%
  as_flex_table() %>%
  theme_vanilla() %>%
  set_caption("Table 0: Baseline characteristics of children (12–23 months), by survey round")

# Print inline
ft_baseline
```

---

## 6. flextable – Format all key results

We convert all previously calculated tibbles into nicely formatted `{flextable}` objects. These will be printed inline and also assembled into a Word report.

```{r flextable-format}
ft_round <- tab_round %>%
  flextable() %>%
  theme_vanilla() %>%
  set_header_labels(
    survey_year = "Round",
    n_clusters  = "Clusters",
    n_children  = "Children",
    zero_dose_unwtd = "Unweighted %",
    zero_dose_wtd   = "Weighted %") %>%
  colformat_double(digits = 1) %>%
  set_caption("Table 1: Sample size and zero-dose prevalence by round")

ft_subgroup <- tab_subgroup %>%
  filter(survey_year == 2019) %>%
  select(variable, group, n_children, zero_dose_wtd) %>%
  flextable() %>%
  theme_vanilla() %>%
  set_header_labels(
    variable = "Stratifier",
    group    = "Level",
    n_children = "N children",
    zero_dose_wtd = "Weighted %") %>%
  colformat_double(digits = 1) %>%
  set_caption("Table 2: Subgroup prevalence (EDHS 2019)")

ft_region <- tab_region %>%
  filter(survey_year == 2019) %>%
  select(dhs_region, n_clusters, n_children, prevalence) %>%
  arrange(desc(prevalence)) %>%
  flextable() %>%
  theme_vanilla() %>%
  set_header_labels(
    dhs_region = "Region",
    n_clusters = "Clusters",
    n_children = "Children",
    prevalence = "Prevalence %"
  ) %>%
  colformat_double(digits = 1) %>%
  set_caption("Table 3: Regional prevalence (EDHS 2019)")

ft_cluster <- tab_cluster_summary %>%
  flextable() %>%
  theme_vanilla() %>%
  set_header_labels(
    survey_year = "Round",
    n_clusters = "Total clusters",
    n_urban = "Urban",
    n_rural = "Rural",
    median_children = "Median children/cluster",
    clusters_zero_n = "Clusters with 0 children"
  ) %>%
  colformat_double(digits = 0) %>%
  set_caption("Table 4: Cluster geolocation summary")

# Direct estimates
ft_national <- national %>%
  flextable() %>%
  theme_vanilla() %>%
  set_header_labels(
    survey_year = "Round",
    prevalence_pct = "Prevalence %",
    se_pct = "SE",
    ci_low = "CI lower",
    ci_high = "CI upper"
  ) %>%
  colformat_double(digits = 1) %>%
  set_caption("Table 5: National direct estimate (95% CI)")

ft_by_region_2019 <- by_region %>%
  filter(survey_year == 2019) %>%
  select(dhs_region, n_clusters, n_children, prevalence_pct, se_pct, ci_low, ci_high) %>%
  arrange(desc(prevalence_pct)) %>%
  flextable() %>%
  theme_vanilla() %>%
  set_header_labels(
    dhs_region = "Region",
    n_clusters = "Clusters",
    n_children = "Children",
    prevalence_pct = "Prevalence %",
    se_pct = "SE",
    ci_low = "CI lower",
    ci_high = "CI upper"
  ) %>%
  colformat_double(digits = 1) %>%
  set_caption("Table 6: Regional direct estimates, EDHS 2019")
```

---

## 7. Print flextables inline

```{r print-tables}
ft_round
ft_cluster
ft_region
ft_subgroup
ft_national
ft_by_region_2019
```

---

## 8. officer – Assemble a complete Word report

Finally, we use `{officer}` to combine **all tables** and **both figures** into a single professionally formatted Word document, saved as `data/clean/table/analysis_report.docx`.

```{r officer-word-report, eval=TRUE}
# Create a new Word document
doc <- read_docx()

# Add title page style
doc <- doc %>%
  body_add_par("Zero-Dose Prevalence Analysis", style = "heading 1") %>%
  body_add_par("EDHS 2011, 2016, 2019", style = "heading 2") %>%
  body_add_par(" ", style = "Normal") %>%
  body_add_par(paste("Generated:", Sys.Date()), style = "Normal")

# Function to add a flextable and a page break
add_ft <- function(doc, ft, caption = NULL) {
  if (!is.null(caption)) {
    doc <- body_add_par(doc, caption, style = "heading 3")
  }
  doc <- body_add_flextable(doc, ft)
  doc <- body_add_par(doc, " ")  # blank line
  doc <- body_add_break(doc)
  doc
}

# Add all tables
doc <- add_ft(doc, ft_baseline, "Baseline characteristics")
doc <- add_ft(doc, ft_round, "Table 1: Round summary")
doc <- add_ft(doc, ft_cluster, "Table 2: Cluster summary")
doc <- add_ft(doc, ft_region, "Table 3: Regional prevalence")
doc <- add_ft(doc, ft_subgroup, "Table 4: Subgroup prevalence")
doc <- add_ft(doc, ft_national, "Table 5: National direct estimates")
doc <- add_ft(doc, ft_by_region_2019, "Table 6: Regional direct estimates (2019)")

# Add figures
doc <- doc %>%
  body_add_par("Figure 1: National trend", style = "heading 3") %>%
  body_add_img(src = "data/clean/figures/direct_est_national_trend.png",
               width = 6, height = 4.5) %>%
  body_add_break() %>%
  body_add_par("Figure 2: Regional estimates (2019)", style = "heading 3") %>%
  body_add_img(src = "data/clean/figures/direct_est_region_2019_ci.png",
               width = 6.5, height = 4.5)

# Save the Word document
print(doc, target = "data/clean/table/analysis_report.docx")

message("\n✅ Word report saved to data/clean/table/analysis_report.docx")
```

---

